Handling String-Number Conversion in String Constraint Solving

### 2019-11-11 draft.tex
try to merge the intro made by Julain and the original one

String constraint solving has received considerable attention in the constraint solving community, with solvers such as CVC4 and Z3 implmenting solvers of ever-increasing sophistication.  Operations such as substring and string length are widely supported, and recent solvers such as Z3-str3 have shown impressive results.  However, these solvers have limited for operations that convert between strings and integers; string length operations are sometimes well supported, but not parsing a string as an integer or turning an integer value into its string form.  When there is support, it is often limited to ground terms, i.e. at least one side of the conversion must be a constant.  Thus, $v = string2int("5")$ or $v = int2string(5)$ are handled but $v_1 = int2string(v_2)$ will fail.

@@

\begin{tabular}{l|c}
	{\tt{x = [0, 1, 2, 3, 4, 5]}} & [0, 1, 2, 3, 4, 5] \\
	{\tt{resetArray(x)}} & [0, 0, 0, 0, 0, 0] \\
	{\tt{x["3"] = 5}} & [0, 0, 0, 5, 0, 0] \\
	{\tt{x["12"] = 7}} & [0, 0, 0, 5, 0, 0, , , , , , 7] \\
	{\tt{resetArray(x)}} & [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0] \\
\end{tabular}

In this code, string and integer values access the same array elements interchangeably, necessitating conversions between string and integers to faithfully model JavaScript semantics.  This conversion is mandated explicitly by JavaScript semantics: the 2019 edition of ECMAScript requires that $ToPropertyKey$ be called on the element expression (\S{12.3.2.1}), and $ToPropertyKey$ calls {\tt{ToString}} on that value in all but special cases (\S{7.1.14}).

Now observe that {\tt{resetArray}} assigns to the array in a loop of which the size cannot, in general, be known a priori; as such, falls into the case where neither the string nor the integer in a conversion are constant and hence most SMT solvers will not be able to handle it.  But this is a rather straightforward array operation, and so this limit will cripple any model of non-trivial JavaScript code.

\todo{The end of Julian's current writing}

\hide{Why this problem is important? Need to say it is used often, e.g., read 
from text file or user input. Why this problem is challenging? It is undecidable. Some tricky cases 
using string as array index. Bounded SAT-based encoding might not work even for 
very simple constraints. e.g., $y=str2int(x) \wedge y>9999999$, if $x$ is a 
ASCII string, then XXX Boolean variables are required. Or we can just propose 
an example that all major solver fails. Maybe we still say our approach is still over-approximation + under-approximation, in this case, need to think how to present the over-approximation part. Maybe no CEGAR is fine.}

@@

Such constraint contains \textit{string-number conversion} functions (e.g., $\sti{input}$ and $\sti{index}$), \textit{equality constraints} (e.g., $index = input\cdot \its{1}$), and \textit{integer constraints} (e.g., $0 \leq \sti{input}<M$). Solving constraint of this kind is a very challenging problem. From the theoretical point of view, this problem is already proven to be undecidable~\cite{day2018satisfiability}. In practice, in our experiments, all of the state-of-the-art string constraint solvers do not support string-number conversion correctly. Some of them handles such constraints with only over-approximation of the original formula, and is known to have one-side error (see Section~\ref{section:evaluation} for detailed explanation). 
In this paper, we propose a framework that handles string constraints with string-number conversion efficiently. Our framework uses both over and under-approximation of the string constraints. The over-approximation is for proving UNSAT and the under-approximation is for proving SAT of the string constraint. Both over and under-approximation should fail in a decidable fragment of string constraints that can be efficiently solved.

>>>>

String constraint solving has received considerable attention in the constraint solving community, with solvers such as CVC4 and Z3 implementing solvers of ever-increasing sophistication.  Operations such as substring and string length are widely supported, and recent solvers such as Z3-str3 have shown impressive results.  However, these solvers have limited for operations that convert between strings and integers; string length operations are sometimes well supported, but not parsing a string as an integer or turning an integer value into its string form.  When there is support, it is often limited to ground terms, i.e. at least one side of the conversion must be a constant.  Thus, $n = \sti{x}$ or $x = \its{5}$ are handled but $n = \sti{x}$ will fail.

@@

\begin{tabular}{l|c}
	{\tt{x = [0, 1, 2, 3, 4, 5]}} & [0, 1, 2, 3, 4, 5] \\
	{\tt{resetArray(x)}} & [0, 0, 0, 0, 0, 0] \\
	{\tt{x[3] = 5}} & [0, 0, 0, 5, 0, 0] \\
	{\tt{x["12"] = 7}} & [0, 0, 0, 5, 0, 0, , , , , , 7] \\
	{\tt{resetArray(x)}} & [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0] \\
\end{tabular}

In this code, string and integer values access the same array elements interchangeably, necessitating conversions between string and integers to faithfully model JavaScript semantics.  This conversion is mandated explicitly by JavaScript semantics: the 2019 edition of ECMAScript requires that $ToPropertyKey$ be called on the element expression (\S{12.3.2.1}), and $ToPropertyKey$ calls {\tt{ToString}} on that value in all but special cases (\S{7.1.14}). On the other hand, the size of an array ({\tt{arr.length}} in the example) is defined as the ``largest integer index plus one'', so the integer value of element expression can affect the array size.

Now observe that {\tt{resetArray}} assigns to the array in a loop of which the size cannot, in general, be known a priori; as such, falls into the case where neither the string nor the integer in a conversion is constant and hence most SMT solvers will not be able to handle it. More specifically, in some program the assignment {\tt{x[k] = 5}} can happen before calling {\tt{resetArray(x)}}, where {\tt{k}} is an user input string. One would need to know the value of $\sti{k}$ and hence whether it will affect the size of the array {\tt{x}}. This is a rather straightforward array operation, and so this limit will cripple any model of non-trivial JavaScript code.

For the purpose of program analysis, string solver also needs to support \emph{equality constraints} (e.g. $x.y = y.x$), \emph{membership constraints} (e.g., $x \in (a.b)^*$), and \emph{integer constraints} (e.g., $|x|-|y|>3$). Most of the standard string operations can be encoded using the above three types of constraints\footnote{Except a few exceptions, such as {\tt{replaceALL}}, {\tt{reverse}}, and string-number conversions.}. Those are also the type of constraints supported by most state-of-the-art string constraint solvers such as Z3, CVC4, Z3Str3, and Trau+.

Indeed, the analysis of string manipulating program is becoming more and more important nowadays. Various infamous security vulnerabilities such as injection and cross-site scripting attack are caused by malicious string input values. In the past, a significant amount of research efforts have been investigated to symbolic execution and test case generation of string manipulating programs~\cite{saxena2010symbolic,artzi2011framework,huang2004securing,sen2013jalangi} and string constraint solving~\cite{kiezun2009hampi,abdulla2014string,zheng2013z3,abdulla2015norn,abdulla2017flatten,wang2016string,abdulla2018trau,chen2019decision,zheng2017z3str2} is the core enabling technology.

\hide{Why this problem is important? Need to say it is used often, e.g., read 
from text file or user input. Why this problem is challenging? It is undecidable. Some tricky cases 
using string as array index. Bounded SAT-based encoding might not work even for 
very simple constraints. e.g., $y=str2int(x) \wedge y>9999999$, if $x$ is a 
ASCII string, then XXX Boolean variables are required. Or we can just propose 
an example that all major solver fails. Maybe we still say our approach is still over-approximation + under-approximation, in this case, need to think how to present the over-approximation part. Maybe no CEGAR is fine.

@@

Such constraint contains \textit{string-number conversion} functions (e.g., $\sti{input}$ and $\sti{index}$), \textit{equality constraints} (e.g., $index = input\cdot \its{1}$), and \textit{integer constraints} (e.g., $0 \leq \sti{input}<M$). }

Solving constraint of this kind is a very challenging problem. From the theoretical point of view, this problem is already proven to be undecidable~\cite{day2018satisfiability}. In practice, in our experiments, all of the state-of-the-art string constraint solvers do not support string-number conversion correctly. Some of them handles such constraints with only over-approximation of the original formula, and is known to have one-side error (see Section~\ref{section:evaluation} for detailed explanation).

In this paper, we propose a framework that handles string constraints with string-number conversion efficiently. Our framework uses both over and under-approximation of the string constraints. The over-approximation is for proving UNSAT and the under-approximation is for proving SAT of the string constraint. Both over and under-approximation should fail in a decidable fragment of string constraints that can be efficiently solved.
